---
title: "Fish Body Size BACI Study"
author: "Eliana Ortega, Gabby Commisso"
date: "2025-07-02"
output:
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    toc_depth: 3
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview
This working script is written to determine if fishes held at the University of New Mexico Museum of Southwestern Biology, dissected for the 2025 Wood Lab CAREER parasite ecology lab work, show any relationship between size, time, and control/impact factor. 


## Initial Steps
First, we load all required packages. 
```{r load_packages}
library(here)
library(tidyverse)
library(ggplot2)
library(lme4)
library(lmerTest)
library(ggeffects)
```

Next, we import the most up-to-date data
```{r data_importing}
#HYBAMA
meta_data<-read.csv(here("data", "all_lots_okay_to_dissect.csv"))
meta_data$combo<-paste(meta_data$CI,meta_data$decade,sep="_")
hyb_ama_today<-read.csv(here("data", "HYBAMA Data_2025.07.02.csv"))
hyb_ama_with_metadata<-merge(hyb_ama_today, meta_data, by = "CatalogNumber", all.x = TRUE)

hybama_today <- hyb_ama_with_metadata %>% 
  janitor::clean_names()
view(hybama_today)

hybama_data <- read.csv(here("data", "/processed/Hybognathus_amarus_processed_human_readable_2025.07.06.csv"))
view(hybama_data)

hybama_data <- hybama_data %>% 
  janitor::clean_names()
view(hybama_data)
```

## Data Exploration
We're interested in seeing if there is a relationship between year collected and fish body size. Let's make a plot to visualize:
```{r data_exploration_sizevstime}
ggplot(data=hybama_data, aes(year_collected,total_length_mm, color=ci)) + 
  geom_point() 
```
## Preliminary Models
Is there any relationship between year and length depending on CI zone?
To figure it out, we'll write a model
```{r prelim_models}
hybama_model <- lmerTest::lmer(total_length_mm~(ci*year_collected) + (1|catalog_number), data= hybama_data)
summary(hybama_model)

#this tests to see if the model well balanced. we want to see a random spread above and below the center line. looks good!
plot(hybama_model)
```

## Plotting Models
Ignore below for now- WIP!
```{r hymaba_plot}
hybama_pred <- ggpredict(hybama_model, c("year_collected [n=100]"))
#plotting
plot(hybama_pred, show_data = TRUE, dot_alpha = 1)+ 
  labs(x = 'Year Collected', y = 'Standard Length in mm',title=NULL)
```