---
title: "Fish Body Size BACI Study"
author: "Eliana Ortega, Gabby Commisso"
date: "2025-07-02"
output:
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    toc_depth: 3
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview
This working script is written to determine if fishes held at the University of New Mexico Museum of Southwestern Biology, dissected for the 2025 Wood Lab CAREER parasite ecology lab work, show any relationship between size, time, and control/impact factor. 


## Initial Steps
First, we load all required packages. 
```{r load_packages}
library(here)
library(tidyverse)
library(ggplot2)
library(lme4)
library(lmerTest)
library(ggeffects)
library(glmmTMB)
library(DHARMa)
library(parameters)
```

Next, we import the most up-to-date data
```{r data_importing}
#METADATA
meta_data<-read.csv(here("data", "all_lots_okay_to_dissect.csv"))
meta_data$combo<-paste(meta_data$CI,meta_data$decade,sep="_")

#HYBAMA
# here is code to read in unpolished dissection data, but since we now have polished data, this part is unnecessary:
hyb_ama_today<-read.csv(here("data", "HYBAMA Data_2025.07.02.csv"))
hyb_ama_with_metadata<-merge(hyb_ama_today, meta_data, by = "CatalogNumber", all.x = TRUE)
hybama_today <- hyb_ama_with_metadata %>% 
  janitor::clean_names()
view(hybama_today)

#here is code to read in polished data:
hybama_data <- read.csv(here("data", "/processed/Hybognathus_amarus_processed_human_readable_2025.07.06.csv"))
view(hybama_data)
hybama_data <- hybama_data %>% 
  janitor::clean_names()
view(hybama_data)

# GAMAFF
# Here, since we don't have polished data yet, I'm importing the most recent dissection data. 
gam_aff_today <- read.csv(here("data", "GAMAFF Dataset_2025.07.14.csv"))
gam_aff_with_metadata<-merge(gam_aff_today, meta_data, by = "CatalogNumber", all.x = TRUE)
gamaff_today <- gam_aff_with_metadata %>% 
  janitor::clean_names()
View(gamaff_today)
```


# HYBOGNATHUS AMARUS

## Data Exploration
We're interested in seeing if there is a relationship between year collected and fish body size. Let's make a plot to visualize:
```{r data_exploration_sizevstime}
#feel free to play around with color (you can enter hex codes for specific values)
ggplot(data=hybama_data, aes(year_collected,total_length_mm, color=ci)) + 
  geom_point() +
  scale_color_manual(values = c("control" = "blue", 
                                "impact" = "red")) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line())
```

## Preliminary Models
Is there any relationship between year and length depending on CI zone?
To figure it out, we'll write a model
```{r prelim_models}
hybama_model <- lmerTest::lmer(total_length_mm~(ci*year_collected) + (1|catalog_number), data= hybama_data)
summary(hybama_model)

#this tests to see if the model well balanced. we want to see a random spread above and below the center line. looks good!
plot(hybama_model)
```

# connor alternate model 
```{r}
# mutate year into a factor
hybama_data <- hybama_data %>% 
  mutate(year_collected = as.factor(year_collected))
print(unique(hybama_data$year_collected))
# levels: 1939 1943 1947 1949 1959 1977 1978 1984 1986 1987 1988 1994 1995 1996 1997 1999 (reference level is 1939)

# how many observations are in each of the years? 
year.reps <- hybama_data %>% 
  group_by(year_collected) %>% 
  summarize(observations = n(), .groups = "drop")

# 1978 has the most repliates, so set this as the reference level
hybama_data <- hybama_data %>%
  mutate(year_collected = relevel(as.factor(year_collected), ref = "1978"))

connor_test <- glmmTMB(total_length_mm ~ ci*year_collected + (1|catalog_number), 
                       data = hybama_data)
simulateResiduals(fittedModel = connor_test, plot = TRUE)
summary(connor_test)
plot(parameters(connor_test))



#### run the same sort of model but using decade instead of year
# add decade column
hybama_data <- hybama_data %>%
  mutate(
    year_collected = as.factor(year_collected),
    decade = as.numeric(as.character(year_collected)) %/% 10 * 10
  )
# convert decade to factor and check levels
hybama_data <- hybama_data %>% 
  mutate(decade = as.factor(decade))
print(unique(hybama_data$decade))
# Levels: 1930 1940 1950 1970 1980 1990

# count up number of replicates in each decade
decade.reps <- hybama_data %>% 
  group_by(decade) %>% 
  summarize(observations = n(), .groups = "drop")
# 1990 has the most reps, so set that as reference 
hybama_data <- hybama_data %>%
  mutate(decade = relevel(as.factor(decade), ref = "1990"))

# now run models

connor_test_decade <- glmmTMB(total_length_mm ~ ci*decade + (1|catalog_number), 
                       data = hybama_data)
simulateResiduals(fittedModel = connor_test_decade, plot = TRUE)
summary(connor_test_decade)
plot(parameters(connor_test_decade))



########### simple model 

connor_test_simple <- glmmTMB(total_length_mm ~ ci + (1|catalog_number) + (1|year_collected), 
                       data = hybama_data)
simulateResiduals(fittedModel = connor_test_simple, plot = TRUE) # variance homogeneous :(
summary(connor_test_simple)
plot(parameters(connor_test_simple))

plot(residuals(connor_test_simple) ~ fitted(connor_test_simple)) # since this dispersion is random it is ok to ignore the significant lavenes test results

# model is a fine fit, no effect of ci is present in fish size


# run with year as fixed effect
connor_test_simple <- glmmTMB(total_length_mm ~ ci + year_collected + (1|catalog_number), 
                       data = hybama_data)
simulateResiduals(fittedModel = connor_test_simple, plot = TRUE) # variance homogeneous :(
summary(connor_test_simple)
plot(parameters(connor_test_simple))

# this model tells you there is no effect of impact BUT there is an effect of certain years, if you want to know if there is a buried effect of certain years DEPENDENT on ci:impact, then you would do an interaction term
```


These model results can be a little trickier to understand, but you can see it shows us a row called "ciimpact:year_collected". This is the slope of the line that represents how the impact category is changing over time. It's a negative slope, but the p-value is insignificant (0.627) so there isn't really much change over time. 

## Plotting Models
I got the plot to work! Now we can use this same format for all the species. 
```{r hymaba_plot}
#the model below will create a line that plots the best fit for how control and impact categories are changing separately through time (interactive effect)
hybama_pred <- ggpredict(hybama_model, c("year_collected [n=100]", "ci"))
#plotting
# i gave this plot the same aesthetics as above: feel free to change!
plot(hybama_pred, show_data = TRUE, dot_alpha = 1) + 
  labs(x = 'Year Collected', y = 'Standard Length in mm',title=NULL) +
    scale_color_manual(values = c("control" = "blue", 
                                "impact" = "red")) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line())
```

# GAMBUSIA AFFINIS

I'm going to turn it over to you! Try to copy what we did for Hybognathus, but edit it for Gambusia. You can steal all my code from above and change what you need. 
- NOTE: the dataframe is already loaded in, and is called "gamaff_today" (see the initial steps section). Some columns may have different names that in the Hybognathus code, so make sure to double check (ex "year_collected" in the hybama_data is called "year_fish_collected" in gamaff_today)

## Data Exploration
```{r}
ggplot(data=gamaff_today, aes(year_fish_collected,total_length_mm, color=ci)) + 
  geom_point() +
  scale_color_manual(values = c("control" = "blue", 
                                "impact" = "red")) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line())
```

## Preliminary Models
```{r}
gamaff_model <- lmerTest::lmer(total_length_mm~(ci*year_fish_collected) + (1|catalog_number), data= gamaff_today)
summary(gamaff_model)
plot(gamaff_model)

# troubleshooting w connor's help
gamaff_model2 <- glmmTMB(total_length_mm~ year_fish_collected + latitude + (1|catalog_number), data= gamaff_today)
summary(gamaff_model2)
# both latitude and year are significant predictors! but ci is not, and neither are interactive effects with year and latitude/ci

gamaff_model3 <- glmmTMB(total_length_mm ~ ci + (1|catalog_number), data=gamaff_today)
summary(gamaff_model3)
```

## Plotting Models
```{r}
gamaff_pred <- ggpredict(gamaff_model2, c("year_fish_collected"))

plot(gamaff_pred, show_data = TRUE, dot_alpha = 1) + 
  labs(x = 'Year Collected', y = 'Standard Length in mm',title=NULL) +
    scale_color_manual(values = c("control" = "blue", 
                                "impact" = "red")) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line())
```

## Testing Connor Models

```{r}

```

