---
title: "Fish Body Size BACI Study"
author: "Eliana Ortega, Gabby Commisso"
date: "2025-07-02"
output:
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    toc_depth: 3
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview
This working script is written to determine if fishes held at the University of New Mexico Museum of Southwestern Biology, dissected for the 2025 Wood Lab CAREER parasite ecology lab work, show any relationship between size, time, and control/impact factor. 


## Initial Steps
First, we load all required packages. 
```{r load_packages}
library(here)
library(tidyverse)
library(ggplot2)
library(lme4)
library(lmerTest)
library(ggeffects)
library(glmmTMB)
library(DHARMa)
library(parameters)
library(stats)
library(parameters)
```

Next, we import the most up-to-date data
```{r data_importing}
# METADATA
meta_data<-read.csv(here("data", "all_lots_okay_to_dissect.csv"))
meta_data$combo<-paste(meta_data$CI,meta_data$decade,sep="_")

# HYBAMA
# here is code to read in unpolished dissection data, but since we now have polished data, this part is unnecessary:
hyb_ama_today<-read.csv(here("data", "HYBAMA Data_2025.08.01.csv"))
hyb_ama_with_metadata<-merge(hyb_ama_today, meta_data, by = "CatalogNumber", all.x = TRUE)
hybama_today <- hyb_ama_with_metadata %>% 
  janitor::clean_names()
view(hybama_today)

# here is code to read in polished data:
hybama_data <- read.csv(here("data", "/processed/Hybognathus_amarus_processed_human_readable_2025.07.06.csv"))
view(hybama_data)
hybama_data <- hybama_data %>% 
  janitor::clean_names()
view(hybama_data)

# GAMAFF
# reading in unpolished data
gam_aff_today<-read.csv(here("data", "GAMAFF Dataset_2025.08.01.csv"))
gam_aff_with_metadata<-merge(gam_aff_today, meta_data, by = "CatalogNumber", all.x = TRUE)
gamaff_today <- gam_aff_with_metadata %>% 
  janitor::clean_names()
view(gamaff_today)

# reading in polished data 
gamaff_data <- read.csv(here("data", "/processed/Gambusia_affinis_processed_human_readable_2025.08.01.csv"))
gamaff_data <- gamaff_data %>% 
  janitor::clean_names()
View(gamaff_data)

# PIMPRO
# reading in unpolished data
pimpro_today <- read.csv(here("data", "PIMPRO Data_2025.08.01.csv"))

#remove 2021+2022 data
pimpro_today <- pimpro_today %>%
  slice(-(1:36))
pimpro_with_metadata<-merge(pimpro_today, meta_data, by = "CatalogNumber", all.x = TRUE)
pimpro_today <- pimpro_with_metadata %>% 
  janitor::clean_names()
view(pimpro_today)
```


# HYBOGNATHUS AMARUS

## Data Exploration
We're interested in seeing if there is a relationship between year collected and fish body size. Let's make a plot to visualize:
```{r data_exploration_sizevstime}
#feel free to play around with color (you can enter hex codes for specific values)
ggplot(data=hybama_data, aes(year_collected,total_length_mm, color=ci)) + 
  geom_point() +
  scale_color_manual(values = c("control" = "#76A973", 
                                "impact" = "#825F87")) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line())
```

## Preliminary Models
Is there any relationship between year and length depending on CI zone?
To figure it out, we'll write a model
```{r prelim_models}
hybama_model <- glmmTMB(total_length_mm~(ci*year_collected) + (1|catalog_number), data= hybama_data)
summary(hybama_model)
# hmm, not super significant! let's try the predictors separately

hybama_model1 <- glmmTMB(total_length_mm~ year_collected + ci + (1|catalog_number), data= hybama_data)
summary(hybama_model1)
# no terms are significant! latitude?

hybama_model2 <- glmmTMB(total_length_mm~ year_collected + latitude + (1|catalog_number), data= hybama_data)
summary(hybama_model2)
#again, nothing! looks like there just isn't a significant change in hybama body size over time
```
## Model Selection

```{r hybama_model_selection}
AIC(hybama_model)
AIC(hybama_model1)
AIC(hybama_model2)
```

## Plotting Models
```{r gamaff_plot}
hybama_pred <- ggpredict(hybama_model1, c("year_collected", "ci"))

plot(hybama_pred, show_data = TRUE, dot_alpha = 1) + 
  labs(x = 'Year Collected', y = 'Standard Length in mm',title=NULL) +
    scale_color_manual(values = c("control" = "#76A973", 
                                "impact" = "#825F87")) +
    scale_fill_manual(values = c("control" = "#76A973", 
                                "impact" = "#825F87")) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line())
```

# connor alternate model (ignore for now) 
```{r}
# mutate year into a factor
hybama_data <- hybama_data %>% 
  mutate(year_collected = as.factor(year_collected))
print(unique(hybama_data$year_collected))
# levels: 1939 1943 1947 1949 1959 1977 1978 1984 1986 1987 1988 1994 1995 1996 1997 1999 (reference level is 1939)

# how many observations are in each of the years? 
year.reps <- hybama_data %>% 
  group_by(year_collected) %>% 
  summarize(observations = n(), .groups = "drop")

# 1978 has the most repliates, so set this as the reference level
hybama_data <- hybama_data %>%
  mutate(year_collected = relevel(as.factor(year_collected), ref = "1978"))

connor_test <- glmmTMB(total_length_mm ~ ci*year_collected + (1|catalog_number), 
                       data = hybama_data)
simulateResiduals(fittedModel = connor_test, plot = TRUE)
summary(connor_test)
plot(parameters(connor_test))



#### run the same sort of model but using decade instead of year
# add decade column
hybama_data <- hybama_data %>%
  mutate(
    year_collected = as.factor(year_collected),
    decade = as.numeric(as.character(year_collected)) %/% 10 * 10
  )
# convert decade to factor and check levels
hybama_data <- hybama_data %>% 
  mutate(decade = as.factor(decade))
print(unique(hybama_data$decade))
# Levels: 1930 1940 1950 1970 1980 1990

# count up number of replicates in each decade
decade.reps <- hybama_data %>% 
  group_by(decade) %>% 
  summarize(observations = n(), .groups = "drop")
# 1990 has the most reps, so set that as reference 
hybama_data <- hybama_data %>%
  mutate(decade = relevel(as.factor(decade), ref = "1990"))

# now run models

connor_test_decade <- glmmTMB(total_length_mm ~ ci*decade + (1|catalog_number), 
                       data = hybama_data)
simulateResiduals(fittedModel = connor_test_decade, plot = TRUE)
summary(connor_test_decade)
plot(parameters(connor_test_decade))



########### simple model 

connor_test_simple <- glmmTMB(total_length_mm ~ ci + (1|catalog_number) + (1|year_collected), 
                       data = hybama_data)
simulateResiduals(fittedModel = connor_test_simple, plot = TRUE) # variance homogeneous :(
summary(connor_test_simple)
plot(parameters(connor_test_simple))

plot(residuals(connor_test_simple) ~ fitted(connor_test_simple)) # since this dispersion is random it is ok to ignore the significant lavenes test results

# model is a fine fit, no effect of ci is present in fish size


# run with year as fixed effect
connor_test_simple <- glmmTMB(total_length_mm ~ ci + year_collected + (1|catalog_number), 
                       data = hybama_data)
simulateResiduals(fittedModel = connor_test_simple, plot = TRUE) # variance homogeneous :(
summary(connor_test_simple)
plot(parameters(connor_test_simple))

# this model tells you there is no effect of impact BUT there is an effect of certain years, if you want to know if there is a buried effect of certain years DEPENDENT on ci:impact, then you would do an interaction term
```


These model results can be a little trickier to understand, but you can see it shows us a row called "ciimpact:year_collected". This is the slope of the line that represents how the impact category is changing over time. It's a negative slope, but the p-value is insignificant (0.627) so there isn't really much change over time. 

## Plotting Models
I got the plot to work! Now we can use this same format for all the species. 
```{r hymaba_plot}
#the model below will create a line that plots the best fit for how control and impact categories are changing separately through time (interactive effect)
hybama_pred <- ggpredict(hybama_model, c("year_collected [n=100]", "ci"))
#plotting
# i gave this plot the same aesthetics as above: feel free to change!
plot(hybama_pred, show_data = TRUE, dot_alpha = 1) + 
  labs(x = 'Year Collected', y = 'Standard Length in mm',title=NULL) +
    scale_color_manual(values = c("control" = "blue", 
                                "impact" = "red")) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line())
```

# GAMBUSIA AFFINIS

I'm going to turn it over to you! Try to copy what we did for Hybognathus, but edit it for Gambusia. You can steal all my code from above and change what you need. 
- NOTE: the dataframe is already loaded in, and is called "gamaff_today" (see the initial steps section). Some columns may have different names that in the Hybognathus code, so make sure to double check (ex "year_collected" in the hybama_data is called "year_fish_collected" in gamaff_today)

## Creating an only randomly selected dataset
```{r gamaff_random}
#first, remove all non-randomly selected fish from the dataset by hand (boo) by manually reading datasheet and removing fish which say they have non-random selection
# using unpolished dataset but the one which includes all fish because we need to see the notes section to know which ones to delete
gamaff_today_random <- gamaff_today %>%
      slice(-c(1, 2, 3, 4, 5, 31, 32, 33, 34, 35, 76, 77, 78, 79, 80, 106, 107, 108, 109, 110, 156, 157, 158, 159, 160, 226, 227, 228, 229, 230, 231, 237, 238, 239, 240, 241, 243, 244, 245, 291, 292, 293, 294, 295, 296, 297, 298, 299, 300, 301, 302, 303, 304, 305, 306, 307, 308, 309, 310, 311, 312, 313, 314, 315, 316, 317, 318, 319, 320, 321, 322, 323, 324, 325, 326, 327, 328, 329, 330, 336, 337, 338, 339, 340, 341, 342, 343, 344, 345, 346, 347, 348, 349, 350))
View(gamaff_today_random)
# ugh that was so annoying. 

# ok now load in the extra data that eliana collected 
gamaff_excess <- read.csv(here("reu/eliana", "gamaff_randomized.csv"))
View(gamaff_excess)
#add metadata
excess_with_metadata<-merge(gamaff_excess, meta_data, by = "CatalogNumber", all.x = TRUE)
excess_with_metadata <- excess_with_metadata %>% 
  janitor::clean_names()
View(excess_with_metadata)

# must make sure columns are consistent in both
gamaff_today_random2 <- gamaff_today_random %>% 
  select(catalog_number, individual_id, fish_species, year_fish_collected, month_fish_collected, day_fish_collected, total_length_mm, standard_length_mm, weight_mg, latitude, longitude, ci, decade, combo)
View(gamaff_today_random2)

excess_with_metadata2 <- excess_with_metadata %>% 
  select(catalog_number, individual_id, fish_species, year_fish_collected, month_fish_collected, day_fish_collected, total_length_mm, standard_length_mm, weight_mg, latitude, longitude, ci, decade, combo)
View(excess_with_metadata2)

#finally... combine!
gamaff_random <- rbind(gamaff_today_random2, excess_with_metadata2)
View(gamaff_random)
```

## Data Exploration
```{r gamaff_explore}
ggplot(data=gamaff_random, aes(year_fish_collected,total_length_mm, color=ci)) + 
  geom_point() +
  scale_color_manual(values = c("control" = "#76A973", 
                                "impact" = "#825F87")) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line())
```

## Preliminary Models
```{r gamaff_models}
gamaff_model <- glmmTMB(total_length_mm~(ci*year_fish_collected) + (1|catalog_number), data= gamaff_random)
summary(gamaff_model)
# no significance whatsoever for this model, which represents our hypothesis (that control fish's body size change differently than impact fish). Let's break it down:

gamaff_model2 <- glmmTMB(total_length_mm~ year_fish_collected + ci +  (1|catalog_number), data= gamaff_random)
summary(gamaff_model2)
# year alone is a moderately significant predictor, but ci is not. what about latitude?

gamaff_model3 <- glmmTMB(total_length_mm~ year_fish_collected + latitude + (1|catalog_number), data= gamaff_random)
summary(gamaff_model3)
# year and latitude are not significant predictors together. what about an interaction?

gamaff_model4 <- glmmTMB(total_length_mm ~ year_fish_collected*latitude +  (1|catalog_number), data=gamaff_random)
summary(gamaff_model4)
#the model can't run the interaction between year collected and latitude, probably because there are too many degrees of freedom
```

## Model Selection

```{r gamaff_model_selection}
AIC(gamaff_model)
AIC(gamaff_model2)
AIC(gamaff_model3)
AIC(gamaff_model4)
#models 2 and 3 are both "best" fits (difference is not significant)
```

## Plotting Models
```{r gamaff_plot}
gamaff_pred <- ggpredict(gamaff_model2, c("year_fish_collected", "ci"))

plot(gamaff_pred, show_data = TRUE, dot_alpha = 1) + 
  labs(x = 'Year Collected', y = 'Standard Length in mm',title=NULL) +
    scale_color_manual(values = c("control" = "#76A973", 
                                "impact" = "#825F87")) +
    scale_fill_manual(values = c("control" = "#76A973", 
                                "impact" = "#825F87")) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line())
```




# ALL TOGETHER

## Combine Data
We have to combine all our datasheets into one single datasheet in order to get information about species!
```{r combo_datasheet}
# first make sure all the datasheets have the same columns: gambusia already has this arrangement since we already did some combining
pimpro_merge <- pimpro_today %>% 
  select(catalog_number, individual_id, fish_species, year_fish_collected, month_fish_collected, day_fish_collected, total_length_mm, standard_length_mm, weight_mg, latitude, longitude, ci, decade, combo)

hybama_merge <- hybama_today %>%
  select(catalog_number, individual_id, fish_species, year_fish_collected, month_fish_collected, day_fish_collected, total_length_mm, standard_length_mm, weight_mg, latitude, longitude, ci, decade, combo)

# now merge them one at a time!
merged1 <- rbind(pimpro_merge, hybama_merge)
merged <- rbind(merged1, gamaff_random)

View(merged)

unique(merged$fish_species)
```

## Models
Now we'll run some models on this merged dataset!
```{r merged_models}
merge_model1 <- glmmTMB(total_length_mm ~ year_fish_collected + fish_species + (1|catalog_number), data=merged)
summary(merge_model1)
#all the fish species are distinct from one another, as noted by their significance codes: that makes sense! let's add in ci and latitude and see if those are important factors

merge_model2 <- glmmTMB(total_length_mm ~ year_fish_collected + fish_species + ci + (1|catalog_number), data=merged)
summary(merge_model2)
# ci isn't important!

merge_model3 <- glmmTMB(total_length_mm ~ year_fish_collected + fish_species + latitude + (1|catalog_number), data=merged)
summary(merge_model3)
# latitude isn't significant either!

# which of these models is best?
AIC(merge_model1)
AIC(merge_model2)
AIC(merge_model3)
#looks like model 3 is the best! we'll go with that: notice it uses latitude as another predictor, which ties in with one of our initial interests in the dataset. 
```

## Plotting Model

Now let's take the best model and make it pretty!
```{r gamaff_plot}
merged_pred <- ggpredict(merge_model3, c("year_fish_collected", "fish_species"))

plot(merged_pred, show_data = TRUE, dot_alpha = 1) + 
  labs(x = 'Year Collected', y = 'Standard Length in mm',title=NULL) +
    scale_color_manual(values = c("Hybognathus amarus" = "#5A86AD", 
                                "Gambusia affinis" = "#825F87", 
                                "Pimephales promelas" = "#C14A09")) +
    scale_fill_manual(values = c("Hybognathus amarus" = "#5A86AD", 
                                "Gambusia affinis" = "#825F87", 
                                "Pimephales promelas" = "#C14A09")) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line())
```

# Explaining Model Output

```{r merge_model3}
summary(merge_model3)
```
You'll notice our model output displays the intercept, year, latitude, and ONLY two species. This is because the model has automatically set the factor that comes first alphabetically (*Gambusia affinis*) as the "reference" factor, or intercept. So what this means: the intercept is 240.213, which means at year 0 (not a real thing in our case), *Gambusia affinis* would be 240mm long, *Hybognathus amarus* would be 240+19.78mm long, and *Pimephales promelas* would be 240+12.28mm long. The significant p-values for HYBAMA and PIMPRO are basically confirming that these are disinct species, and isn't super relevant for us. 

Instead, we're more interested in the year and latitude factors. These estimates are both NEGATIVE, which means that as years increase, size decreases; and as latitude increases, size decreases. However, neither of them are significant, so we can't really interpret them. Essentially, none of these species are changing their body size through time. 