---
title: "ABQ Daily Report"
author: "Connor Whalen, Gabby Commisso, Katie Leslie, Chelsea Wood"
date: "2025-06-10"
output:
  html_document:
    theme: readable
    highlight: textmate
    toc: true
    toc_float: true
    toc_depth: 3
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
# Description
Summary log of leadership and decision activities conducted by leadership team at the University of New Mexico during our 2025 summer CAREER parasite lab work in Albuquerque, New Mexico. 

# Initial Selection Procedure

### Link to Google Spreadsheet
[Lot selection and tracking spreadsheet](https://docs.google.com/spreadsheets/d/1waDMf6cTmtEeaT_14-uCFXRKaHarc8dxyh9Y7yJx9l0/edit?pli=1&gid=486468529#gid=486468529)

### 6/11/25
Connor sent resampling lots and numbers to Chelsea who replaced them in the spreadsheet on 6/12/25. The lots are as follows:
MSB:Fish:5253	Gambusia affinis
MSB:Fish:5247	Hybognathus amarus
MSB:Fish:5803	Hybognathus amarus
MSB:Fish:11156 Pimephales promelas
MSB:Fish:11167	Pimephales promelas
MSB:Fish:1171	Hybognathus amarus
MSB:Fish:1150	Hybognathus amarus
MSB:Fish:3651	Pimephales promelas
MSB:Fish:30386	Gambusia affinis
MSB:Fish:30239	Pimephales promelas
MSB:Fish:19232	Gambusia affinis
MSB:Fish:73766	Hybognathus amarus
MSB:Fish:77030	Hybognathus amarus
MSB:Fish:77021	Hybognathus amarus
MSB:Fish:94735	Pimephales promelas

### 6/12/25
Connor found two more lots that need to be resampled. Went back through the code file specimen_tallies.R and checked for resampling options for both GAMAFF 1068 and HYBAMA 1162. Neither of them are able to be replaced as there are no more available lots in the collection for their respective CI category and years. Connor made note of this in the spreadsheet The lots are as follows:
MSB:Fish:1068	Gambusia affinis
MSB:Fish:1162	Hybognathus amarus

### 6/13/25
Connor found more lots that needed to be resampled. Connor also went through the list of missing lots and has identified the majority of those that were not found initially, though some are still missing. We have 1 lot that needs to be either found or resampled to send in the shipment to Katie, as well as a few other lots that are missing. The list of missing lots is here:
MSB:Fish:7917 Pimephales promelas
MSB:Fish:18794 Pimephales promelas
MSB:Fish:334 Gambusia affinis
MSB:Fish:1122	Hybognathus amarus
MSB:Fish:24944 Hybognathus amarus
MSB:Fish:39103 Hybognathus amarus


### Check at the outset to make sure that we are on track to reach our goals with the lots that we have in hand
Install required packages
```{r, eval=FALSE, echo=FALSE}
install.packages("here")
```

Load required packages
```{r, echo=FALSE}
library(googledrive)
library(skimr)
library(lubridate)
library(tidyverse)
library(here)
```

First, bring in all of the meta-data you need to unite with your dissection data
```{r}
meta_data <- read.csv(here("data", "all_lots_okay_to_dissect.csv"))
meta_data$combo<-paste(meta_data$CI,meta_data$decade,sep="_")
```


Then bring in the dataset
```{r}
lots<-read.csv(here("data", "final_lots_2024.12.30_WITH_SHIP_REQUEST_missing_lots_replaced.csv"))
length(lots$CatalogNumber)

lots_in_hand <- lots %>%
filter(ready!= "SKIP")

length(lots_in_hand$CatalogNumber)
219-203
```

Now merge lot data with meta-data (all.x makes sure that you keep all individual fish from the same lot, even though they have the same catalog number).
```{r}
lots_with_metadata<-merge(lots_in_hand, meta_data, by.x = "CatalogNumber", by.y = "CatalogNumber", all.x = TRUE)
length(lots_with_metadata$CatalogNumber)
```

Plot to see where the dissected fish fall
```{r}
plot(jitter(lots_with_metadata$Latitude.y,30)~jitter(lots_with_metadata$YearCollected.y,5))
```

Actual level of replication
```{r}
lots_actual<-lots_with_metadata %>%
  group_by(ScientificName.x,combo.x) %>%
  summarize(actual = n())

View(lots_actual)

count_up<-lots_with_metadata %>%
  group_by(ScientificName.x) %>%
  summarize(actual = 5*n())
```

Tally up everything that is not being shipped
```{r}
dissect_here<-lots_with_metadata %>%
filter(SHIP_REQUEST != "x") %>%
  group_by(ScientificName.x) %>%
  summarize(actual = 5*n())
  
sum(dissect_here$actual)
```

I checked lots_actual against the database, and it looks like we have pulled everything it is possible to pull!
The only chance of filling in gaps is finding uncatalogued specimens. I'll take a quick look and see if any of the old jars are by chance not cataloged.
- No uncatalogued lots to be found!


### Code for doing daily checks

The purpose of this script is to allow dissectors to quickly assess levels of replication in each category (i.e., before-impact, after-impact, before-control, after-control) during dissections, so that we can adjust lot selection on the fly if need be. We also want to make sure that fish body size range is constant across decades, that latitudes sampled in early days are the same as the latitudes sampled later, and that we are on track to hit our dissection goals.


### First, bring in all of the meta-data you need to unite with your dissection data
```{r}
meta_data<-read.csv(here("data", "all_lots_okay_to_dissect.csv"))
meta_data$combo<-paste(meta_data$CI,meta_data$decade,sep="_")
View(meta_data)
```


# HYBAMA
Download latest file 
```{r}
hyb_ama_today<-read.csv(here("data", "HYBAMA Data_2025.07.03.csv"))
                        
length(hyb_ama_today$CatalogNumber)
```

Now merge dissection data with meta-data (all.x makes sure that you keep all individual fish from the same lot, even though they have the same catalog number).
```{r}
hyb_ama_with_metadata<-merge(hyb_ama_today, meta_data, by = "CatalogNumber", all.x = TRUE)
length(hyb_ama_with_metadata$CatalogNumber)

View(hyb_ama_with_metadata)
```

Plot to see where the dissected fish fall
```{r}
plot(jitter(hyb_ama_with_metadata$Latitude,30)~jitter(hyb_ama_with_metadata$YearCollected,5))+abline(a = 35.211106, b = 0, lty = 2)+abline(v = 1953, lty = 2)
```


Desired level of replication
```{r}
hyb_ama_desired<-read.csv(here("data", "hyb_ama_goal.csv"))
hyb_ama_desired<-hyb_ama_desired[,-1]
colnames(hyb_ama_desired)[2]<-"n_lots_desired"
hyb_ama_desired$individual_fish_desired<-hyb_ama_desired$n_lots_desired*5
```


Actual level of replication
```{r}
hyb_ama_actual<-hyb_ama_with_metadata %>%
  group_by(combo) %>%
  summarize(actual = n())
```


Merge desired and actual
```{r}
hyb_ama_remaining<-merge(hyb_ama_desired, hyb_ama_actual, by.x = "combo", by.y = "combo", all.x = TRUE, all.y = TRUE)
```


Calculate how many you have left
```{r}
hyb_ama_remaining$to_go<-(hyb_ama_remaining$individual_fish_desired-ifelse(is.na(hyb_ama_remaining$actual),0,hyb_ama_remaining$actual))
sum(hyb_ama_remaining$actual,na.rm=T)
sum(hyb_ama_remaining$to_go,na.rm=T)
```

Check that body size is invariant through time
```{r}
summary(glm(hyb_ama_with_metadata$Total_Length_mm~hyb_ama_with_metadata$YearCollected))
plot(hyb_ama_with_metadata$Total_Length_mm~hyb_ama_with_metadata$YearCollected)
summary(glm(hyb_ama_with_metadata$Total_Length_mm~hyb_ama_with_metadata$Latitude))
plot(hyb_ama_with_metadata$Total_Length_mm~hyb_ama_with_metadata$Latitude)
```

Create a matrix that shows how many fish you are finishing per day
```{r}
hyb_ama_dailies<-hyb_ama_with_metadata %>%
  group_by(Completion_Date) %>%
  summarize(actual = n())
```


Need to resample a lot of hybama 6/24/25:
Lot 5272 does not have big enough fish, and there are no remaining lots for control 1981-1990, so we will skip this lot. 

### 6/30/25
Replace HYBAMA MSB:Fish:18820 -- impact_1991to2000: replace with MSB:Fish:19001 --- too small, replace with MSB:Fish:30075

### 7/2/25
Replace HYBAMA MSB:Fish:30401 (too few fish in the jar) -- impact_1991to2000 replace with MSB:Fish:20390

# GAMAFF
Download latest file 
```{r}
gam_aff_today<-read.csv(here("data", "GAMAFF Dataset_2025.07.17.csv"))
                        
length(gam_aff_today$CatalogNumber)
```

Now merge dissection data with meta-data (all.x makes sure that you keep all individual fish from the same lot, even though they have the same catalog number).
```{r}
gam_aff_with_metadata<-merge(gam_aff_today, meta_data, by = "CatalogNumber", all.x = TRUE)
length(gam_aff_with_metadata$CatalogNumber)

View(gam_aff_with_metadata)
```

Plot to see where the dissected fish fall
```{r}
plot(jitter(gam_aff_with_metadata$Latitude,30)~jitter(gam_aff_with_metadata$YearCollected,5))+abline(a = 35.211106, b = 0, lty = 2)+abline(v = 1953, lty = 2)
```


Desired level of replication
```{r}
gam_aff_desired<-read.csv(here("data", "gam_aff_goal.csv"))
gam_aff_desired<-gam_aff_desired[,-1]
colnames(gam_aff_desired)[2]<-"n_lots_desired"
gam_aff_desired$individual_fish_desired<-gam_aff_desired$n_lots_desired*5
```


Actual level of replication
```{r}
gam_aff_actual<-gam_aff_with_metadata %>%
  group_by(combo) %>%
  summarize(actual = n())
```


Merge desired and actual
```{r}
gam_aff_remaining<-merge(gam_aff_desired, gam_aff_actual, by.x = "combo", by.y = "combo", all.x = TRUE, all.y = TRUE)
```


Calculate how many you have left
```{r}
gam_aff_remaining$to_go<-(gam_aff_remaining$individual_fish_desired-ifelse(is.na(gam_aff_remaining$actual),0,gam_aff_remaining$actual))
sum(gam_aff_remaining$actual,na.rm=T)
sum(gam_aff_remaining$to_go,na.rm=T)
```

Check that body size is invariant through time
```{r}
summary(glm(gam_aff_with_metadata$Total_Length_mm~gam_aff_with_metadata$YearCollected))
plot(gam_aff_with_metadata$Total_Length_mm~gam_aff_with_metadata$YearCollected)
summary(glm(gam_aff_with_metadata$Total_Length_mm~gam_aff_with_metadata$Latitude))
plot(gam_aff_with_metadata$Total_Length_mm~gam_aff_with_metadata$Latitude)
```

Create a matrix that shows how many fish you are finishing per day
```{r}
gam_aff_dailies<-gam_aff_with_metadata %>%
  group_by(Completion_Date) %>%
  summarize(actual = n())
```

### 7/6/25
Today we began GAMAFF!

### 7/9/25
MSB:Fish:1077 has a bunch/majority of fish in it that are cut open. Randomly reselected MSB:Fish:1105 (see specimen_tallies.R)

### 7/15/25
Ran the check with yesterday's data. On Friday 7/10 we implemented a size-by-year rule, and on Monday 7/14 we implemented a size-by-latitude rule. Both relationships are still significant. More details about these selection criteria can be found in the methods document. 

### 7/16/25
Today we uncovered MSB:Fish:4440, where there the fish are too small to fit the latitude rule. We checked alternative lots to replace, but found that we are already dissecting every lot in this category except for one which was deemed too small by Connor (MSB:Fish:3993). So we went ahead with dissections on 4440, even though they don't fit the new latitude size rule. 

We replaced MSB:Fish:1080 (only 10 fish) with MSB:Fish:1089. 

MSB:Fish:340 has only ten fish in the jar, but there is not another lot from the same decade/impact combo. Initially we skipped this jar, but later Chelsea opted to replace it. It was collected in 1949 so I attempted to replace it with a lot from "impact_1951-1960", but apparently there is only one jar in that category and we already dissected it. 

### 7/18/25
Discovered a lot (MSB:Fish:7239) where there are not 5 fish that fit the latitude rule. HOWEVER, as of 7/17/25, our latitude trend has disappeared. So, Gabby made the executive call to go ahead and dissect the jar anyways, making it a TRUE random selection. This was noted on the datasheet. 

### 7/21/25
Discovered a lot (MSB:Fish:6300) where there are not 5 fish that fit the latitude rule. We decided to go ahead and dissect the jar anyways since the latitude trend has disappeared. This was noted on the datasheet. 

# PIMPRO

Download latest file 
```{r}
pim_pro_today<-read.csv(here("data", "PIMPRO Data_2025.07.28.csv"))
                        
length(pim_pro_today$CatalogNumber)
```

Remove fish dissected prior to being at UNM
```{r}
View(pim_pro_today)
pim_pro_today <- pim_pro_today %>%
  slice(-(1:36))
```

Now merge dissection data with meta-data (all.x makes sure that you keep all individual fish from the same lot, even though they have the same catalog number).
```{r}
pim_pro_with_metadata<-merge(pim_pro_today, meta_data, by = "CatalogNumber", all.x = TRUE)
length(pim_pro_with_metadata$CatalogNumber)

View(pim_pro_with_metadata)
```

Plot to see where the dissected fish fall
```{r}
plot(jitter(pim_pro_with_metadata$Latitude,30)~jitter(pim_pro_with_metadata$YearCollected,5))+abline(a = 35.211106, b = 0, lty = 2)+abline(v = 1953, lty = 2)
```


Desired level of replication
```{r}
pim_pro_desired<-read.csv(here("data", "pim_pro_goal.csv"))
pim_pro_desired<-pim_pro_desired[,-1]
colnames(pim_pro_desired)[2]<-"n_lots_desired"
pim_pro_desired$individual_fish_desired<-pim_pro_desired$n_lots_desired*5
```


Actual level of replication
```{r}
pim_pro_actual<-pim_pro_with_metadata %>%
  group_by(combo) %>%
  summarize(actual = n())
```


Merge desired and actual
```{r}
pim_pro_remaining<-merge(pim_pro_desired, pim_pro_actual, by.x = "combo", by.y = "combo", all.x = TRUE, all.y = TRUE)
```


Calculate how many you have left
```{r}
pim_pro_remaining$to_go<-(pim_pro_remaining$individual_fish_desired-ifelse(is.na(pim_pro_remaining$actual),0,pim_pro_remaining$actual))
sum(pim_pro_remaining$actual,na.rm=T)
sum(pim_pro_remaining$to_go,na.rm=T)
```

Check that body size is invariant through time
```{r}
summary(glm(pim_pro_with_metadata$Total_Length_mm~pim_pro_with_metadata$YearCollected))
plot(pim_pro_with_metadata$Total_Length_mm~pim_pro_with_metadata$YearCollected)
summary(glm(pim_pro_with_metadata$Total_Length_mm~pim_pro_with_metadata$Latitude))
plot(pim_pro_with_metadata$Total_Length_mm~pim_pro_with_metadata$Latitude)
```

Create a matrix that shows how many fish you are finishing per day
```{r}
pim_pro_dailies<-pim_pro_with_metadata %>%
  group_by(Completion_Date) %>%
  summarize(actual = n())
```