---
title: "Processing Data from Raw to Analyzable Form"
author: "Chelsea Wood, Gabby Commisso, Connor Whalen"
date: "June-August 2025"
output:
  html_document:
  theme: readable
highlight: textmate
toc: true
toc_float: true
toc_depth: 3
---

The purpose of this script is to clean the raw dataset, strip out unneeded fields, sum parasite counts across organs, and generally get the thing into analyzable form. If you ever need things from the raw dataset, they will still be there for you to access!  But the processed dataset will have what is needed for most analyses.


First step: load required packages
```{r packages}
library(googledrive)
library(skimr)
library(lubridate)
library(tidyverse)
library(reshape2)
library(readxl)
library(readr)
library(dplyr)
library(here)
```


Here's the workflow that you need to follow for each fish species:
1. List all parasite species detected and figure out which are valid (i.e., lumping some, discarding those that aren't parasites).
2. Sum across organs within valid parasite species and double counts for those organs that need to be doubled.
3. Create a dataset containing only the data you want (get rid of organ and voucher headers).
4. Melt the dataset so that parasite counts are in one column and parasite names are in another column.


First, bring in all of the meta-data you need to unite with your dissection data

```{r metadata}
meta_data <- read.csv(here("data", "all_lots_okay_to_dissect.csv"))
meta_data$combo<-paste(meta_data$CI,meta_data$decade,sep="_")
```


# HYBAMA (*Hybognathus amarus*)

Download most recent file
```{r hybama_dataset}
hybama_today<-read.csv(here("data", "HYBAMA data_2025.07.03.csv"))
length(hybama_today$CatalogNumber)
```

Now merge dissection data with meta-data (all.x makes sure that you keep all individual fish from the same lot, even though they have the same catalog number).

```{r hybama_merge}
hybama_with_metadata<-merge(hybama_today, meta_data, by.x = "CatalogNumber", by.y = "CatalogNumber", all.x = TRUE)
length(hybama_with_metadata$CatalogNumber)
```
Let's make this dataset tidy! Making headers consistent, removing unnecessary columns, etc
```{r tidy_hybama}
hybama_with_metadata <- hybama_with_metadata %>% 
  janitor::clean_names()

hybama_with_metadata_tidy <- hybama_with_metadata %>% 
  select(-c(73:132))

hybama_with_metadata_tidy <- hybama_with_metadata_tidy %>% 
  select (-c(fish_species_1, fish_barcode_1, msb_fish_catalog_number, tissues, island, island_group, verbatim_elevation, lat_long_comments))

view(hybama_with_metadata_tidy)
```

Plot to see where the dissected fish fall.
```{r hybama_test}
plot(jitter(hybama_with_metadata_tidy$latitude,30)~jitter(hybama_with_metadata$year_collected,5))+abline(a = 35.211106, b = 0, lty = 2)+abline(v = 1953, lty = 2)

str(hybama_with_metadata_tidy)

hybama_headers<-ls(hybama_with_metadata_tidy)
hybama_headers<-as.factor(hybama_headers)
```

## Add parasites across organs and double what needs to be doubled.
Notes:
- cyst.unk is NOT a parasite
- trem_unk_conn and trem_unk_flush are columns of entirely zeros, so were not included in this
- myxo.susp was determined to be NOT a parasite
```{r hybama_parasites}
mono.gyro<-2*(hybama_with_metadata_tidy$mono_gyro_pectoralfin)+2*(hybama_with_metadata_tidy$mono_gyro_pelvicfin)+hybama_with_metadata_tidy$mono_gyro_analfin+hybama_with_metadata_tidy$mono_gyro_caudalfin+hybama_with_metadata_tidy$mono_gyro_dorsalfin+2*(hybama_with_metadata_tidy$mono_gyro_gill)+hybama_with_metadata_tidy$mono_gyro_flush

trem.fim <- 2*(hybama_with_metadata_tidy$trem_fim_pectoralfin)+hybama_with_metadata_tidy$trem_fim_caudalfin

cope.lern <- 2*(hybama_with_metadata_tidy$cope_lern_pelvicfin)+hybama_with_metadata_tidy$cope_lern_analfin+hybama_with_metadata_tidy$cope_lern_flush

cope.imler <- hybama_with_metadata_tidy$cope_imler_analfin

trem.meta.unk <- hybama_with_metadata_tidy$trem_meta_unk_caudalfin + hybama_with_metadata_tidy$trem_meta_unk_caudalfin + 2*(hybama_with_metadata_tidy$meta_unk_gill)

trem.diplo <- hybama_with_metadata_tidy$trem_diplo_liver + 2*(hybama_with_metadata_tidy$trem_diplo_eye) + hybama_with_metadata_tidy$trem_diplo_conn + hybama_with_metadata_tidy$trem_diplo_flush

trem.d <- hybama_with_metadata_tidy$trem_d_liver + hybama_with_metadata_tidy$trem_d_conn + hybama_with_metadata_tidy$trem_d_flush

trem.gold <- hybama_with_metadata_tidy$trem_gold_liver + hybama_with_metadata_tidy$trem_gold_conn

trem.em <- hybama_with_metadata_tidy$trem_em_eye 

trem.l <- hybama_with_metadata_tidy$trem_l_conn + hybama_with_metadata_tidy$trem_l_flush

trem.b <- hybama_with_metadata_tidy$trem_b_conn + hybama_with_metadata_tidy$trem_b_flush

trem.dlum <- hybama_with_metadata_tidy$trem_dlum_flush

trem.ridge <- hybama_with_metadata_tidy$trem_ridge_flush

nem.cl <- hybama_with_metadata_tidy$nem_cl_int

nem.unk <- hybama_with_metadata_tidy$nema_unk_int

mono.dact <- 2*(hybama_with_metadata_tidy$mono_dact_gill)

myxo.b <- 2*(hybama_with_metadata_tidy$myxo_b_gill)
```

Quickly evaluating relationship between total length and weight to check for any potential abnormalities.
```{r hybama_weight_check}
ggplot(hybama_with_metadata_tidy,aes(total_length_mm,weight_mg))+
  geom_point(size=4)
```
Looks good!

## Putting it all together

Put the columns into one dataframe...
```{r hybama_all_together_now}
hybama_processed_data<-cbind.data.frame(
  hybama_with_metadata_tidy$catalog_number,
  hybama_with_metadata_tidy$year_fish_collected,        
  hybama_with_metadata_tidy$month_collected, 
  hybama_with_metadata_tidy$day_fish_collected,         
  hybama_with_metadata_tidy$individual_id,
  hybama_with_metadata_tidy$fish_barcode,
  hybama_with_metadata_tidy$dissector,
  hybama_with_metadata_tidy$examiner, 
  hybama_with_metadata_tidy$dissection_date, 
  hybama_with_metadata_tidy$sex, 
  hybama_with_metadata_tidy$total_length_mm, 
  hybama_with_metadata_tidy$standard_length_mm,
  hybama_with_metadata_tidy$weight_mg, 
  hybama_with_metadata_tidy$ci, 
  hybama_with_metadata_tidy$combo, 
  hybama_with_metadata_tidy$latitude, 
  hybama_with_metadata_tidy$longitude, 
  cope.imler, 
  cope.lern, 
  mono.dact, 
  mono.gyro, 
  myxo.b, 
  nem.cl, 
  nem.unk, 
  trem.b, 
  trem.d, 
  trem.diplo, 
  trem.dlum, 
  trem.em, 
  trem.fim, 
  trem.gold, 
  trem.l, 
  trem.meta.unk, 
  trem.ridge)

```


Name the columns...
```{r hybama_columns}
colnames(hybama_processed_data)[1]<-"CatalogNumber"
colnames(hybama_processed_data)[2]<-"YearCollected"
colnames(hybama_processed_data)[3]<-"MonthCollected"
colnames(hybama_processed_data)[4]<-"DayCollected"
colnames(hybama_processed_data)[5]<-"IndividualFishID"
colnames(hybama_processed_data)[6] <- "FishBarcode"
colnames(hybama_processed_data)[7]<-"Dissector"
colnames(hybama_processed_data)[8] <- "Examiner"
colnames(hybama_processed_data)[9]<-"DissectionDate"
colnames(hybama_processed_data)[10]<-"Sex"
colnames(hybama_processed_data)[11]<-"TotalLength_mm"
colnames(hybama_processed_data)[12]<-"StandardLength_mm"
colnames(hybama_processed_data)[13]<-"Weight_mg"
colnames(hybama_processed_data)[14]<-"CI"
colnames(hybama_processed_data)[15]<-"combo"
colnames(hybama_processed_data)[16]<-"Latitude"
colnames(hybama_processed_data)[17]<-"Longitude"

view(hybama_processed_data)
```

## Make the dataset analyzable

```{r make_dataset_analyzable}
hybama_processed_data_longer<-melt(hybama_processed_data,id=c(
  "CatalogNumber", 
  "YearCollected", 
  "MonthCollected", 
  "DayCollected", 
  "IndividualFishID", 
  "FishBarcode", 
  "Dissector",
  "Examiner", 
  "DissectionDate", 
  "Sex", 
  "TotalLength_mm", 
  "StandardLength_mm",
  "Weight_mg",
  "CI","combo",
  "Latitude",
  "Longitude"))

colnames(hybama_processed_data_longer)[18]<-"psite_spp"
colnames(hybama_processed_data_longer)[19]<-"psite_count"

# Fix data type for weight. Otherwise, it is read as a character and weight is plotted wrong.

hybama_processed_data_longer$Weight_mg <- as.numeric(hybama_processed_data_longer$Weight_mg)
```

## Export both sheets

```{r hybama_export}
write.csv(hybama_processed_data_longer, file= "Hybognathus_amarus_processed_machine_readable_2025.07.06.csv")

write.csv(hybama_processed_data, file="Hybognathus_amarus_processed_human_readable_2025.07.06.csv")
```

# GAMAFF (*Gambusia affinis*)

Download most recent file
```{r gamaff_dataset}
gamaff_today<-read.csv(here("data", "GAMAFF Dataset_2025.08.01.csv"))
length(gamaff_today$CatalogNumber)
```

Now merge dissection data with meta-data (all.x makes sure that you keep all individual fish from the same lot, even though they have the same catalog number).

```{r gamaff_merge}
gamaff_with_metadata<-merge(gamaff_today, meta_data, by.x = "CatalogNumber", by.y = "CatalogNumber", all.x = TRUE)
length(gamaff_with_metadata$CatalogNumber)
```
Let's make this dataset tidy! Making headers consistent, removing unnecessary columns, etc
```{r tidy_gamaff}
gamaff_with_metadata <- gamaff_with_metadata %>% 
  janitor::clean_names()
View(gamaff_with_metadata)

#gamaff_with_metadata_tidy <- gamaff_with_metadata %>% 
 # select(-c(73:132))

#hybama_with_metadata_tidy <- hybama_with_metadata_tidy %>% 
# select (-c(fish_species_1, fish_barcode_1, msb_fish_catalog_number, tissues, island, island_group, verbatim_elevation, lat_long_comments))

#view(gamaff_with_metadata_tidy)
```

Plot to see where the dissected fish fall.
```{r gamaff_test}
plot(jitter(gamaff_with_metadata$latitude,30)~jitter(gamaff_with_metadata$year_collected,5))+abline(a = 35.211106, b = 0, lty = 2)+abline(v = 1953, lty = 2)

str(gamaff_with_metadata)

gamaff_headers<-ls(gamaff_with_metadata)
gamaff_headers<-as.factor(gamaff_headers)
```

## Add parasites across organs and double what needs to be doubled.
Notes:
- cyst.unk is NOT a parasite
- trem.unk IS a parasite, since even though it is unidentified, it was at least recognized as a trematode
- nem.sty is not a parasite: determined to be free-living 
- cest.four is not a parasite
- after conferring with Irvin and seeing multiple specimens, I believe cest.a and cest.botsp are the same parasite, and a is just a more mature form. I have combined them in the below code
```{r gamaff_parasites}
crus.d <- gamaff_with_metadata$crus_d_skin + 2*(gamaff_with_metadata$crus_d_gill)

crus.lersp <- gamaff_with_metadata$crus_lersp_flush

mono.gyrof <- gamaff_with_metadata$mono_gyro_analfin + gamaff_with_metadata$mono_gyro_dorsalfin

mono.gyrog <- gamaff_with_metadata$mono_gyro_gill

mono.salsp <- 2*(gamaff_with_metadata$mono_salsp_gill)

mono.ss <- 2*(gamaff_with_metadata$mono_ss_gills)

trem.dips <- 2*(gamaff_with_metadata$trem_dips_eye) + gamaff_with_metadata$trem_dips_connective_tissue + gamaff_with_metadata$trem_dips_flush

trem.iz <- gamaff_with_metadata$trem_iz_intestine

trem.unk <- gamaff_with_metadata$trem_unk_gill

nem.larv <- gamaff_with_metadata$nem_larv_connectivetissue + gamaff_with_metadata$nema_larv_intestine

nem.l <- gamaff_with_metadata$nema_l_connective_tissue

nem.myst <- gamaff_with_metadata$nema_myst_intestine

nem.sc <- gamaff_with_metadata$nema_sc_intestine

nem.unk <- gamaff_with_metadata$nema_unk_intestine

acanth.spk <- gamaff_with_metadata$acanth_spk_flush

cest.botsp <- gamaff_with_metadata$cest_a_intestine + gamaff_with_metadata$cest_botsp_intestine

myxo.myxid <- gamaff_with_metadata$myxo_mxyid_gallbladder_presence1_absence0

myxo.g <- 2*(gamaff_with_metadata$myxo_g_gill)

prot.cil <- gamaff_with_metadata$prot_cil_gill_presence_absence
```

Quickly evaluating relationship between total length and weight to check for any potential abnormalities.
```{r gamaff_weight_check}
ggplot(gamaff_with_metadata,aes(total_length_mm,weight_mg))+
  geom_point(size=4)
```
Looks good!

## Putting it all together

Put the columns into one dataframe...
```{r gamaff_all_together_now}
gamaff_processed_data<-cbind.data.frame(
  gamaff_with_metadata$catalog_number,
  gamaff_with_metadata$year_fish_collected,        
  gamaff_with_metadata$month_collected, 
  gamaff_with_metadata$day_fish_collected,         
  gamaff_with_metadata$individual_id,
  gamaff_with_metadata$fish_barcode,
  gamaff_with_metadata$dissector,
  gamaff_with_metadata$examiner, 
  gamaff_with_metadata$dissection_date, 
  gamaff_with_metadata$sex, 
  gamaff_with_metadata$total_length_mm, 
  gamaff_with_metadata$standard_length_mm,
  gamaff_with_metadata$weight_mg, 
  gamaff_with_metadata$ci, 
  gamaff_with_metadata$combo, 
  gamaff_with_metadata$latitude, 
  gamaff_with_metadata$longitude, 
  crus.d, 
  crus.lersp, 
  mono.gyrof,
  mono.salsp, 
  mono.ss,
  trem.dips, 
  trem.iz, 
  trem.unk, 
  nem.larv, 
  nem.l, 
  nem.myst,
  acanth.spk, 
  cest.botsp, 
  myxo.myxid, 
  myxo.g)
```


Name the columns...
```{r gamaff_columns}
colnames(gamaff_processed_data)[1]<-"CatalogNumber"
colnames(gamaff_processed_data)[2]<-"YearCollected"
colnames(gamaff_processed_data)[3]<-"MonthCollected"
colnames(gamaff_processed_data)[4]<-"DayCollected"
colnames(gamaff_processed_data)[5]<-"IndividualFishID"
colnames(gamaff_processed_data)[6] <- "FishBarcode"
colnames(gamaff_processed_data)[7]<-"Dissector"
colnames(gamaff_processed_data)[8] <- "Examiner"
colnames(gamaff_processed_data)[9]<-"DissectionDate"
colnames(gamaff_processed_data)[10]<-"Sex"
colnames(gamaff_processed_data)[11]<-"TotalLength_mm"
colnames(gamaff_processed_data)[12]<-"StandardLength_mm"
colnames(gamaff_processed_data)[13]<-"Weight_mg"
colnames(gamaff_processed_data)[14]<-"CI"
colnames(gamaff_processed_data)[15]<-"combo"
colnames(gamaff_processed_data)[16]<-"Latitude"
colnames(gamaff_processed_data)[17]<-"Longitude"

view(gamaff_processed_data)
```

## Make the dataset analyzable

```{r gamaff_analyzable}
gamaff_processed_data_longer<-melt(gamaff_processed_data,id=c(
  "CatalogNumber", 
  "YearCollected", 
  "MonthCollected", 
  "DayCollected", 
  "IndividualFishID", 
  "FishBarcode", 
  "Dissector",
  "Examiner", 
  "DissectionDate", 
  "Sex", 
  "TotalLength_mm", 
  "StandardLength_mm",
  "Weight_mg",
  "CI","combo",
  "Latitude",
  "Longitude"))

colnames(gamaff_processed_data_longer)[18]<-"psite_spp"
colnames(gamaff_processed_data_longer)[19]<-"psite_count"

# Fix data type for weight. Otherwise, it is read as a character and weight is plotted wrong.

gamaff_processed_data_longer$Weight_mg <- as.numeric(gamaff_processed_data_longer$Weight_mg)
```

## Export both sheets

```{r gamaff_export}
write.csv(gamaff_processed_data_longer, file= "Gambusia_affinis_processed_machine_readable_2025.08.01.csv")

write.csv(gamaff_processed_data, file="Gambusia_affinis_processed_human_readable_2025.08.01.csv")
```

# PIMPRO (*Pimephales promelas*)

Download most recent file
```{r pimpro_dataset}
pimpro_today<-read.csv(here("data", "PIMPRO Data_08.12.2025.csv"))
length(pimpro_today$CatalogNumber)
```

Now merge dissection data with meta-data (all.x makes sure that you keep all individual fish from the same lot, even though they have the same catalog number).

```{r pimpro_merge}
pimpro_with_metadata<-merge(pimpro_today, meta_data, by.x = "CatalogNumber", by.y = "CatalogNumber", all.x = TRUE)
length(pimpro_with_metadata$CatalogNumber)

# remove fishes collected in 2021 or 2022
pimpro_with_metadata <- pimpro_with_metadata %>%
  filter(year_fish_collected != "2021") %>%
  filter(year_fish_collected != "2022") %>%
  filter(year_fish_collected != "2025")
View(pimpro_with_metadata)
```
Let's make this dataset tidy! Making headers consistent, removing unnecessary columns, etc
```{r tidy_pimpro}
pimpro_with_metadata <- pimpro_with_metadata %>% 
  janitor::clean_names()
View(pimpro_with_metadata)

```

Plot to see where the dissected fish fall.
```{r pimpro_test}
plot(jitter(pimpro_with_metadata$latitude,30)~jitter(pimpro_with_metadata$year_collected,5))+abline(a = 35.211106, b = 0, lty = 2)+abline(v = 1953, lty = 2)

str(pimpro_with_metadata)

pimpro_headers<-ls(pimpro_with_metadata)
pimpro_headers<-as.factor(pimpro_headers)
colnames(unique(pimpro_with_metadata))

```

## Add parasites across organs and double what needs to be doubled.
Notes:
- unk_caudalfin may be a parasite but we don't know what, so we are dropping it 
- cyst_unkn_stomach may be a parasite but we don't know what, so we are dropping it 
_ cyst_unkn_intestine may be a parasite but we don't know what, so we are dropping it 
- unk_gallbladder_presence_absence may be a parasite but we don't know what, so we are dropping it
```{r pimpro_parasites}
trem.meta <- pimpro_with_metadata$trem_meta_skin + pimpro_with_metadata$trem_meta_muscle

crus.lercyp <- pimpro_with_metadata$crus_lercyp_skin + pimpro_with_metadata$crus_fin_dorsalfin + pimpro_with_metadata$crus_lercyp_pelvicfin

trem.diphur <- pimpro_with_metadata$trem_diphur_muscle + pimpro_with_metadata$trem_diphur_liver + pimpro_with_metadata$trem_diphur_connective_tissue + pimpro_with_metadata$trem_diphur_flush + pimpro_with_metadata$trem_diphur_intestine

trem.n <- pimpro_with_metadata$trem_n_body_cavity + pimpro_with_metadata$trem_n_connectivetissue 

mono.gyrsp <- pimpro_with_metadata$mono_gyrsp_dorsalfin + 2*(pimpro_with_metadata$mono_gyrsp_gill) + pimpro_with_metadata$mono_gyrsp_pelvicfin + 2*(pimpro_with_metadata$mono_gyrsp_pectoralfin)

mono.unk <- 2*(pimpro_with_metadata$mono_unkn_pectoralfin)

trem.white <- pimpro_with_metadata$trem_white_liver + pimpro_with_metadata$trem_white_conn

trem.p <- pimpro_with_metadata$trem_p_liver + pimpro_with_metadata$trem_p_connective_tissue + pimpro_with_metadata$trem_p_flush + pimpro_with_metadata$trem_p_kidney

trem.neasp <- 2*(pimpro_with_metadata$trem_neasp_eye)

trem.unk <- pimpro_with_metadata$trem_unk_connectivetissue + pimpro_with_metadata$trem_unk_flush

trem.acet <- pimpro_with_metadata$trem_acet_connectivetissue

trem.unclear <- pimpro_with_metadata$trem_unclear_conn

cest.larv <- pimpro_with_metadata$cest_larv_connective_tissue

nem.cont <- pimpro_with_metadata$nem_cont_connectivetissue + pimpro_with_metadata$nem_cont_intestine

trem.cent <- pimpro_with_metadata$trem_cent_flush + pimpro_with_metadata$trem_cent_intestine + pimpro_with_metadata$trem_acet_connectivetissue

nem.cap <- pimpro_with_metadata$nem_cap_intestine

cest.b <- pimpro_with_metadata$cest_b_intestine

trem.buc <- pimpro_with_metadata$trem_buc_intestine

nem.unk <- pimpro_with_metadata$nem_unk_intestine

nem.larv <- pimpro_with_metadata$nem_larv_intestine + pimpro_with_metadata$nem_larv_gallbladder

myxo.myxsp <- pimpro_with_metadata$myxo_myxsp_gill_number_of_clusters

myxo.m <- 2*(pimpro_with_metadata$myxo_m_gill)

mono.dacsp <- 2*(pimpro_with_metadata$mono_dacsp_gill)

```

Quickly evaluating relationship between total length and weight to check for any potential abnormalities.
```{r pimpro_weight_check}
ggplot(pimpro_with_metadata,aes(total_length_mm,weight_mg))+
  geom_point(size=4)

```


## Putting it all together

Put the columns into one dataframe...
```{r pimpro_all_together_now}
pimpro_processed_data<-cbind.data.frame(
  pimpro_with_metadata$catalog_number,
  pimpro_with_metadata$year_fish_collected,        
  pimpro_with_metadata$month_collected, 
  pimpro_with_metadata$day_fish_collected,         
  pimpro_with_metadata$individual_id,
  pimpro_with_metadata$fish_barcode,
  pimpro_with_metadata$dissector,
  pimpro_with_metadata$examiner, 
  pimpro_with_metadata$dissection_date, 
  pimpro_with_metadata$sex, 
  pimpro_with_metadata$total_length_mm, 
  pimpro_with_metadata$standard_length_mm,
  pimpro_with_metadata$weight_mg, 
  pimpro_with_metadata$ci, 
  pimpro_with_metadata$combo, 
  pimpro_with_metadata$latitude, 
  pimpro_with_metadata$longitude, 
  trem.meta,
  crus.lercyp,
  trem.diphur,
  trem.n,
  mono.gyrsp,
  mono.unk,
  trem.white,
  trem.p,
  trem.neasp,
  trem.unk,
  trem.acet, 
  trem.unclear,
  cest.larv,
  nem.cont,
  trem.cent,
  nem.cap,
  cest.b,
  trem.buc,
  nem.unk,
  nem.larv,
  myxo.myxsp,
  myxo.m,
  mono.dacsp
  )
```


Name the columns...
```{r pimpro_columns}
colnames(pimpro_processed_data)[1]<-"CatalogNumber"
colnames(pimpro_processed_data)[2]<-"YearCollected"
colnames(pimpro_processed_data)[3]<-"MonthCollected"
colnames(pimpro_processed_data)[4]<-"DayCollected"
colnames(pimpro_processed_data)[5]<-"IndividualFishID"
colnames(pimpro_processed_data)[6] <- "FishBarcode"
colnames(pimpro_processed_data)[7]<-"Dissector"
colnames(pimpro_processed_data)[8] <- "Examiner"
colnames(pimpro_processed_data)[9]<-"DissectionDate"
colnames(pimpro_processed_data)[10]<-"Sex"
colnames(pimpro_processed_data)[11]<-"TotalLength_mm"
colnames(pimpro_processed_data)[12]<-"StandardLength_mm"
colnames(pimpro_processed_data)[13]<-"Weight_mg"
colnames(pimpro_processed_data)[14]<-"CI"
colnames(pimpro_processed_data)[15]<-"combo"
colnames(pimpro_processed_data)[16]<-"Latitude"
colnames(pimpro_processed_data)[17]<-"Longitude"

view(pimpro_processed_data)
```

## Make the dataset analyzable

```{r pimpro_analyzable}
pimpro_processed_data_longer<-melt(pimpro_processed_data,id=c(
  "CatalogNumber", 
  "YearCollected", 
  "MonthCollected", 
  "DayCollected", 
  "IndividualFishID", 
  "FishBarcode", 
  "Dissector",
  "Examiner", 
  "DissectionDate", 
  "Sex", 
  "TotalLength_mm", 
  "StandardLength_mm",
  "Weight_mg",
  "CI","combo",
  "Latitude",
  "Longitude"))

colnames(pimpro_processed_data_longer)[18]<-"psite_spp"
colnames(pimpro_processed_data_longer)[19]<-"psite_count"

# Fix data type for weight. Otherwise, it is read as a character and weight is plotted wrong.

pimpro_processed_data_longer$Weight_mg <- as.numeric(pimpro_processed_data_longer$Weight_mg)
```

## Export both sheets

```{r pimpro_export}
write.csv(pimpro_processed_data_longer, file= "Pimephales_promelas_processed_machine_readable_2025.08.14.csv")

write.csv(pimpro_processed_data, file="Pimephales_promelas_processed_human_readable_2025.08.14.csv")
getwd()
```

## Merge all data into one file
```{r}
all.fish <- pimpro_processed_data %>%
  full_join(gamaff_processed_data, by = "CatalogNumber") %>%
  full_join(hybama_processed_data, by = "CatalogNumber")
```

